Sending a Balloon to the Edge of Space

### 2018-11-14 paper/paper.rmd
continue work on paper. some mac stuff didnt work for johns graphs, redo for linux?

```{r setup, include=FALSE}
library(knitr)
library(mosaic)
library(gridExtra)
opts_chunk$set(echo = F)

@@

There are four main layers to the atmosphere: the troposphere, stratosphere, mesosphere, and thermosphere. The **troposphere** is the section of the atmosphere from roughly 0km to 15km, and is the layer in which we live. It includes 75% of the mass of all gases in the atmosphere, consisting of 78% nitrogen gas, 21% oxygen, and less than 1% of argon, carbon dioxide, and other constituents. Because it is both the lowest in altitude and most massive layer, the pressure in the troposphere is the greatest: 100kPa. As you increase altitude in the troposphere, temperature decreases to about -60C. The **stratosphere** is the section from roughly 15km to 50km. The ozone layer, or the region in the atmosphere which absorbs most UV rays, is within the stratosphere at about 20km in altitude. The pressure in the stratosphere is 10kPa. The **mesosphere** is the region from 50km to 80km. The pressure in the mesosphere is 0.1 kPa. Finally, the **thermosphere** is the highest in altitude of the four layers, spanning from 80km into outer space. Here, the pressure is only 0.01kPa, as it is the least dense of the layers. The few air molecules in the thermosphere are heated to extremely high temperatures by the Sun’s radiation. However, because there are so few molecules, we cannot measure an appreciable difference in overall temperature, so the effective temperature of the thermosphere is less than that of the mesosphere.

@@

*shortly before launch it was discovered that the Geiger Counter was non-operable and therefore it was taken out of the payload. Instead, Simon Oros focused on ____.

@@

The Vernier Gas Pressure Sensor, model GPS-BTA, was calibrated to the Extech SD700 Barometric Pressure/Humidity/Temperature Datalogger, which was used as our standard for pressure. The Vernier sensor was connected to an Arduino microcontroller with a Vernier Analog Protoboard Adaptor and configured to log data onto a micro-SD card every 1.50 seconds. Instead of having a output unit of hPa or kPa, the Arduino maps input voltages between 0 to 5 volts to 10-bit integer values between 0 and 1023—a total of 1024 bins with 0.49 mV resolution. Thus, each unit represents 4.9 mV. The SD700 has a pressure range of 10 to 1100 hPa, fine resolution of 0.1 hPa, optimal temperature range of 0 to 50°C, response time of 10 milliseconds, and total accuracy with its factory calibration ±4 kPa. It requires 6 AAA batteries, and records data to an external SD card in the format of an Excel worksheet. It was configured so that it sampled data every five seconds to have smoother and more gradual changes in pressure. After both sensors were turned on at the same time to easily compare the data later, they were immediately placed in a vacuum chamber, in which air and other gasses are removed by a vacuum pump, thus creating a low-pressure environment inside the chamber. The release valve was closed and the vacuum pump was opened. When the pressure reading from the valve reached its minimum of approximately 0 hPa, the vacuum pump was then disconnected. An important part of the pressure calibration process was having “plateaus,” or multiple data points at various pressure levels. To have plateaus in the data, the release valve was slowly opened and closed right after, and this process was repeated after waiting approximately 45 seconds multiple times. Graph 5 and 6 show the varying pressure readings from the standard Extech sensor and Arduino Vernier sensor across time, respectively.

```{r vernier_calibration, echo=F}
th <- theme(plot.title = element_text(size=10))

@@

Graph 7, shown below, shows the conversion between the output of the standard (Extech) sensor and that of the Arduino (Vernier) sensor, including the equation to convert the Arduino’s 0.49 mV resolution units to kPa.

```{r vernier_cal2, fig.height=3.5}
pressure3 <- read.csv("data/pressure3.csv")
names(pressure3)[1] <- "Vernier sensor (10-bit unit)"
names(pressure3)[2] <- "Extech sensor (kPa)"

lm_eqn_p3 <- function(df){
    m <- lm(df$`Extech sensor (kPa)` ~ df$`Vernier sensor (10-bit unit)`, df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(coef(m)[1], digits = 3), 
              b = format(coef(m)[2], digits = 3), 
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));                 
}

p3 <- ggplot(pressure3, aes(x=pressure3$`Vernier sensor (10-bit unit)`, y=pressure3$`Extech sensor (kPa)`)) + geom_point() + labs(title="Graph 7: Pressure conversion between\n standard (Extech) and Arduino (Vernier) sensor", x="Pressure from Vernier (10-bit unit)", y="Pressure from Extech (kPa)") + geom_smooth(method = "lm") + geom_text(x=300, y=90, label=lm_eqn_p3(pressure3), parse=TRUE)
p3

@@

humidity1 <- read.csv("data/humidity1.csv")
lm_eqn_h1 <- function(df){
    m <- lm(df$Arduino.... ~ df$Standard...., df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(coef(m)[1], digits = 3), 
              b = format(coef(m)[2], digits = 3), 
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));                 
}
h1 <- ggplot(humidity1, aes(x=humidity1$Arduino...., y=humidity1$Standard....)) + geom_point() + labs(title="Graph 8: Humidity conversion between\n standard (Vernier) and Arduino (BME280) sensor", x="Relative humidity from BME280 (% RH)", y="Relative humidity from Vernier (% RH)") + geom_smooth(method = "lm") + geom_text(x=55, y=60, label=lm_eqn_h1(humidity1), parse=TRUE)
h1
```

@@

```{r bme_temp, fig.height=4}
temp1 <- read.csv("data/temp1.csv")
lm_eqn_t1 <- function(df){
    m <- lm(df$Arduino...C. ~ df$Fieldpiece...C., df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(coef(m)[1], digits = 3), 
              b = format(coef(m)[2], digits = 3), 
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));                 
}
t1 <- ggplot(temp1, aes(x=temp1$Arduino...C., y=temp1$Fieldpiece...C.)) + geom_point() + labs(title="Graph 9: Temperature conversion between\n standard (Fieldpiece) and Arduino (BME280) sensor", x="Temperature from BME280 (°C)", y="Temperature from Fieldpiece's T1 rod (°C)") + geom_smooth(method = "lm") + geom_text(x=5, y=20, label=lm_eqn_t1(temp1), parse=TRUE)
t1
errors <- read.csv("data/errors.csv")
names(errors)[2] <- "Percent error (%)"
```

>>>>

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
library(latex2exp)
opts_chunk$set(echo = F, warning = F)

@@

There are four main layers to the atmosphere: the troposphere, stratosphere, mesosphere, and thermosphere. The **troposphere** is the section of the atmosphere from roughly 0km to 15km, and is the layer in which we live. It includes 75% of the mass of all gases in the atmosphere, consisting of 78% nitrogen gas, 21% oxygen, and less than 1% of argon, carbon dioxide, and other constituents. Because it is both the lowest in altitude and most massive layer, the pressure in the troposphere is the greatest: 100kPa. As altitude increases in the troposphere, temperature decreases to about -60C. The** stratosphere** is the section from roughly 15km to 50km. The ozone layer, or the region in the atmosphere which absorbs most UV rays, is within the stratosphere at about 20km in altitude. The pressure in the stratosphere is 10kPa. The mesosphere is the region from 50km to 80km. The pressure in the mesosphere is 0.1 kPa. Finally, the thermosphere is the highest in altitude of the four layers, spanning from 80km into outer space. Here, the pressure is only 0.01kPa, as it is the least dense of the layers. The few air molecules in the thermosphere are heated to extremely high temperatures by the Sun’s radiation. However, because there are so few molecules, we cannot measure an appreciable difference in overall temperature, so the effective temperature of the thermosphere is less than that of the mesosphere.

@@

*shortly before launch it was discovered that the Geiger Counter was non-operable and therefore it was taken out of the payload. Instead, Simon Oros focused on ground-level radiation measurements.

@@

The Vernier Gas Pressure Sensor, model GPS-BTA, was calibrated to the Extech SD700 Barometric Pressure/Humidity/Temperature Datalogger, which was used as our standard for pressure. The Vernier sensor was connected to an Arduino microcontroller with a Vernier Analog Protoboard Adaptor and configured to log data onto a micro-SD card every 1.50 seconds. Instead of having a output unit of hPa or kPa, the Arduino maps input voltages between 0 to 5 volts to 10-bit integer values between 0 and 1023—a total of 1024 bins with 0.0049 V resolution. Thus, each unit represents 4.9 mV. The SD700 has a pressure range of 10 to 1100 hPa, fine resolution of 0.1 hPa, optimal temperature range of 0 to 50°C, response time of 10 milliseconds, and total accuracy with its factory calibration ±4 kPa. It requires 6 AAA batteries, and records data to an external SD card in the format of an Excel worksheet. It was configured so that it sampled data every five seconds to have smoother and more gradual changes in pressure. After both sensors were turned on at the same time to easily compare the data later, they were immediately placed in a vacuum chamber, in which air and other gasses are removed by a vacuum pump, thus creating a low-pressure environment inside the chamber. The release valve was closed and the vacuum pump was opened. When the pressure reading from the valve reached its minimum of approximately 0 hPa, the vacuum pump was then disconnected. An important part of the pressure calibration process was having “plateaus,” or multiple data points at various pressure levels. To have plateaus in the data, the release valve was slowly opened and closed right after, and this process was repeated after waiting approximately 45 seconds multiple times. Graph 5 and 6 show the varying pressure readings from the standard Extech sensor and Arduino Vernier sensor across time, respectively.

```{r vernier_calibration, echo=F, fig.align='left'}
th <- theme(plot.title = element_text(size=10))

@@

Graph 7, shown below, shows the conversion between the output of the standard (Extech) sensor and that of the Arduino (Vernier) sensor, including the equation to convert the Arduino’s 4.9 mV resolution units to kPa.

&nbsp;

```{r vernier_cal2, fig.height=3.5, warning=F}
pressure3 <- read.csv("data/pressure3.csv")
names(pressure3)[1] <- "Vernier sensor (10-bit unit)"
names(pressure3)[2] <- "Extech sensor (kPa)"

p3 <- ggplot(pressure3, aes(x=pressure3$`Vernier sensor (10-bit unit)`, y=pressure3$`Extech sensor (kPa)`)) + geom_point() + labs(title="Graph 7: Pressure conversion between\n standard (Extech) and Arduino (Vernier) sensor", x="Pressure from Vernier (10-bit unit)", y="Pressure from Extech (kPa)") + geom_smooth(method = "lm") + geom_text(x=300, y=90, label=TeX("y = -25.3 + 0.255x, r^2 = 1"), parse=TRUE)

p3

@@

humidity1 <- read.csv("data/humidity1.csv")
lm_eqn_h1 <- function(df){
    m <- lm(df$Arduino.... ~ df$Standard...., df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(coef(m)[1], digits = 3), 
              b = format(coef(m)[2], digits = 3), 
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));                 
}
h1 <- ggplot(humidity1, aes(x=humidity1$Arduino...., y=humidity1$Standard....)) + geom_point() + labs(title="Graph 8: Humidity conversion between\n standard (Vernier) and Arduino (BME280) sensor", x="Relative humidity from BME280 (% RH)", y="Relative humidity from Vernier (% RH)") + geom_smooth(method = "lm") + geom_text(x=55, y=60, label=TeX("y = 9.12 + 1.07x, r^2 = 0.997"), parse=TRUE)
h1
```

@@

&nbsp;

```{r bme_temp, fig.height=4}
temp1 <- read.csv("data/temp1.csv")
lm_eqn_t1 <- function(df){
    m <- lm(df$Arduino...C. ~ df$Fieldpiece...C., df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(coef(m)[1], digits = 3), 
              b = format(coef(m)[2], digits = 3), 
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));                 
}
t1 <- ggplot(temp1, aes(x=temp1$Arduino...C., y=temp1$Fieldpiece...C.)) + geom_point() + labs(title="Graph 9: Temperature conversion between\n standard (Fieldpiece) and Arduino (BME280) sensor", x="Temperature from BME280 (°C)", y="Temperature from Fieldpiece's T1 rod (°C)") + geom_smooth(method = "lm") + geom_text(x=5, y=20, label=TeX("y = 0.122 + 1.03x, r^2 = 0.999"), parse=TRUE)
t1
errors <- read.csv("data/errors.csv")
names(errors)[2] <- "Percent error (%)"
```
